\LARGE \bf Human-Robot Interaction: A Deliberative Architecture

### 2013-07-11 paper.tex
Fixed many typos/style issues

\documentclass{elsarticle}

\usepackage{cite}
\usepackage{graphicx, subfigure}
%\usepackage{hyperref}
\usepackage{url}
\graphicspath{{figs/}}
\DeclareGraphicsExtensions{.pdf,.jpg,.png}

\usepackage{ifthen}

@@

\newcommand{\stmt}[1]{{\footnotesize \tt $\langle$ #1\relax$\rangle$}}
%\newcommand{\stmt}[1]{{\footnotesize $\langle\stmttt#1\relax\rangle$}}
%\newcommand{\rawstmt}[1]{{\footnotesize \stmttt#1\relax}}
%\def\stmttt#1 #2 #3\relax{
%\texttt{\IfBeginWith{#1}{?}{\textbf{#1}}{#1} \IfBeginWith{#2}{?}{\textbf{#2}}{\emph{#2}} \ifthenelse {\equal{#3}{true} \OR \equal{#3}{false}}{\emph{#3}}{\IfBeginWith{#3}{?}{\textbf{#3}} {#3}}}}

%\newcommand{\setstmt}[1]{{\footnotesize [\setstmttt#1\relax]}}
%\def\setstmttt#1,#2\relax{\stmttt#1\relax, \stmttt#2\relax}

@@

Over the last years we have focused our efforts on identifying the cognitive
prerequisites of these challenges, and giving them experimental reality on the
robots: what is required for sentences like ``Let's dress the table together''
to be understood by the robot, correctly interpreted in the spatial and
temporal context of the interaction, and eventually transformed into a set of
actions.

We have chosen to tackle the challenge from several ends: human-aware
navigation and motion planning (\cite{Mainprice2011}), situation assessment
coupled with motion planning (\cite{Mainprice2012}), projection of
``mightabilities'' that anticipates what surrounding agents may do
(\cite{Pandey2011}), and the design and deployment of a pervasive
\emph{knowledge-centric} software architecture.

The last item is the focus of this paper: what does ``knowledge-centric''
architecture means, and what are its benefits and drawbacks.

@@

In the next section, we first introduce the deliberative layer of our robots at
the functional level. It represents what we call a \emph{knowledge-oriented}
architecture, built around a central knowledge management tool called the {\sc
Oro} server.

The following sections go over each of the facets of this deliberative layer:
how knowledge is produced asynchronously from geometric reasoning, how the
robot conducts grounded multi-modal interaction with humans, how knowledge is
used by decision-making components like the robot controller or the task
planner, and finally, how our knowledge-centric architecture enables the
implementation of specific internal cognitive processes.

@@

\subsection{The LAAS Deliberative Layer}

Fig.~\ref{fig|archi} give an overview of the connections between the
deliberative components of our architecture~\cite{Alami2011}.

@@

This architecture moves away from standard layered approaches. Interactions
between components are mostly bidirectional and, from the software components
point of view, we do not introduce layers of abstraction (we do, however, have
access to the lower level modules of the robot to execute actions, but all
cognition-related modules reside at the same level). This is especially visible
for the dialogue input processing. This component does not simply act as an
alternative perceptual input to the symbolic database, but also actively
queries previously acquired knowledge to disambiguate and validate the newly
created symbolic knowledge (see section~\ref{sect|com}).

Our architecture relates but is to be distinguished from \emph{Beliefs,
Desires, Intentions} (BDI) architectures. BDI architectures are primarily
focused on \emph{practical reasoning}, \ie the process of deciding, step by
step, which action to perform to reach a goal (as summarised by
Woolridge~\cite{Woolridge1999}). The management of the interaction between
knowledge (the beliefs) and task and plan representation and execution (the
desires and the intentions) is central, and aims at selecting at each step the
best subgoal. It becomes then an intention that the robot commits to.

This interaction between knowledge and actions is also central to our approach
(as for any cognitive system), but task representation and task execution is
not seen as a monolithic, central function: it is one of the activities of the
robot, actually split between communication components (that can acquire
desires from interaction with agents, amongst other things) and an execution
controller that may decide to take an incoming desire into account to create
its own internal goals. The controller generates and controls intentions from
these goals with the help of a symbolic task planner, that has also direct
access to the knowledge base.

The architecture is not only focused on this workload, and other
activities are conducted without being explicitly considered as desires:
assessment of the situation and the environment, dialogue (including
performative dialogue that can possibly change the internal state of the robot,
but does not lead to the creation of desires,  like question answering or
statement assertion), various background monitoring and recognition tasks, etc.

Regarding the anchoring question, this architecture is bidirectional. The
components we described provide a \textit{bottom-up} grounding process:
geometric reasoning and dialogue processing modules constantly build and push
new symbolic contents about the world to the knowledge base where it becomes
accessible to decisional layers. In parallel, the knowledge base relies on
reasoning in a \textit{top-down} way to produce new facts that may trigger in
return physical behaviours.

@@

In our architecture (Fig.~\ref{fig|archi}), knowledge manipulation relies on a
\emph{semantic blackboard}: one central server (the {\sc Oro}
server~\cite{Lemaignan2010}) stores knowledge as it is produced by each of the
deliberative components. It conversely exposes an API to query the knowledge
base.

@@

Relying on RDF triples and Description Logics has major advantages like the
availability of numerous open-source libraries to manipulate the ontology,
interoperability with several major on-line knowledge bases (like {\sc
OpenCyc}, {\sc WordNet} or {\sc DBPedia}), open-world reasoning, and the formal
guarantee of decidability (it is always possible to classify a Description
Logics ontology).

It has also notable limitations: RDF triples imply only binary predicates
(\stmt{subject predicate object}), which constrains the expressiveness of the
system or leads to cumbersome reifications. Alternative exists (like {\sc
KnowRob}~\cite{Tenorth2009a}) that mixes RDF with more expressive logic
languages like {\sc Prolog}, at the price, however, of other limitations, like
closed-world reasoning or immutable T-Box. The classification
performance is another issue: from our experience, with an ontology sized for a
standard experiment (about 100 classes and 200 instances), classification
typically take about 100ms, which becomes problematic during interactions.
Besides, the performances are difficult to predict, since a seemingly
inoffensive new statement may indirectly change radically the logical
complexity of the whole knowledge model and lead to notable degradation of
classification time.

This knowledge model also largely exclude representation of continuous
phenomena (like time) or uncertain phenomena. When required, these have to be
managed by dedicated components.

Amongst the features of {\sc Oro}, the system can dynamically creates
new independent knowledge models (\ie independent instantiations of the
ontology). This feature is used for instance to store separately the beliefs of
each of the agents the robot interacts with (see section~\ref{sect|tom}).

@@

Theory of Mind (originally defined in~\cite{Premack1978}) is the cognitive
ability that a subject possesses to represent the mental state of another
agent, possibly including knowledge that contradicts the subject own model: for
example, a book can be at the same time \emph{visible} by myself, and \emph{not
visible} by you.

This skill, which is essential to understand others' perspectives during
interactions, appears in children around their third year. It supposes the ability to build, store and retrieve separate models of the knowledge of the interactors.

@@

One classical application of this cognitive skills is the so-called
\emph{False-Belief} experiment (also known as the \emph{Sally and Ann}
experiment)~\cite{Leslie2000}: a child is asked to watch a scene where two
people, A and B, manipulate objects. Then A leaves and B hides away one
object. When A comes back, we ask the child ``where do you think A will
look for the object?''. Before acquiring a theory of mind, children are not
able to separate their own (true) model of the world (where they know that
the object was hidden) from the model of A, which contains \emph{false
beliefs} on the world (A still thinks the object is at its original
position since he did not see B hiding it). Using seperate knowledge models
in the knowledge base, we have been able to replicate this experience on
our robots~\cite{Warnier2012a}.

@@

The {\sc Oro} server also feature a mechanism to mimic simple forms of
biological memory.  When new statements are inserted in the knowledge base, a
\emph{memory profile} is optionally attached to them.

@@

We rely however on this short term memory for a particular use-case:
\emph{active concept}. Some modules, like the natural language processor, use
the {\tt short term} memory profile to mark for a few seconds important
concepts that are currently manipulated by the robot. For example, if a human
asks the robot: ``Give me all red objects'', the human, the \concept{Give}
action, and each red objects that are found are successively marked as
\emph{active concepts} by inserting statements like \stmt{human type
ActiveConcept} in the short-term memory (which can be considered, in this case,
to be a working memory). We use this feature to give a (visual) feedback to the
users.

@@

Anchoring perceptions in a symbolic model requires perception abilities and
their symbolic interpretation. We rely on a dedicated geometric and temporal
reasoning module (as named on Fig.~\ref{fig|archi}) called SPARK (\emph{SPAtial
Reasoning \& Knowledge}~\cite{Sisbot2011}). It is a situation assessment
reasoner that generates symbolic knowledge from the geometry of the
environment with respect to relations between objects, robots and humans,
also taking into account the different perspective that each agent has on
the environment (Fig.~\ref{fig|spark-oro}).

SPARK is an \emph{amodal} geometric model of the environment that serves both
as basis for the fusion of the perception modalities and as bridge with the
symbolic layer. This geometric model is continuously updated at run-time by the
robot based on its sensors. In our experiments, objects are identified and
localised through 2D barcodes.  On the other hand, the human pose is tracked
with a Kinect-like device (assisted by motion capture to accurately track the
head motion, which is required to compute what the human is looking at).

@@

This allows us to deal with ambiguities that arises when one speaker refers to
an object within a reference system (or changes the reference system, \ie
switches perspective) without making the reference frame
explicit~\cite{Breazeal2006, Ros2010}.

As a result, the robot stores models on the environment either in the
\emph{ego-centric} reference frame (from the robot perspective) or in
addressee-centred (\ie, \emph{allo-centric}) frames.

@@

Natural language processing is one of the fields of human-robot interaction
where the introduction of the semantic layer has been most beneficial.
We detail in~\cite{Lemaignan2011a} the techniques and the tool called
\texttt{dialogs} that we have developed for English language parsing and
grounding, along with verbalisation and (minimalist) dialogue management.

Natural language input (in experiments, we rely on an Android-based interface,
with Google speech recognition) is parsed into a grammatical structure, and
each sentence atoms are resolved with the help of the ontology to ground
concepts like objects (\ie when a user says ``pick the can'', resolve to which
instance of \emph{can} the user is referring to) and actions. Sentences are
sorted into questions, desires and statements, and processed accordingly.

The system supports quantification (``give me \{a | the | some | all | any | ...\} can''),
thematic roles (action-specific predicates that qualify the actions), interactive disambiguation (the robot asks
questions when it needs more information), anaphora resolution (``give \emph{it} to me'') based on
dialogue history. It also permits knowledge extension by learning new semantic
structures (for instance, a sentence like ``learn that cats are animals'' is
converted into \stmt{Cat subClassOf Animal}, interprets common temporal and place adverbs (like \emph{above} or \emph{tomorrow}) and translates to a certain extend \emph{states} (``I'm tired'') into
\emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1 hasFeature
tired}).

@@

Besides, even knowledge that belongs the \emph{common knowledge} may not be
appropriate in a given interaction context. For instance, the system may
compute that at a given instant the human is looking at the object: \stmt{human
looksAt obj1}. This property makes sense to both parties, but in the context of
the \emph{spygame}, we would like to mainly use immanent properties, not
volatile like a gaze. More research is required to identify relevant
interaction contexts and knowledge classes attached to them.

@@

Because all components rely on the same RDF formalism to format their outputs,
the different communication modalities (\emph{explicit} like verbal, deictic or
based on gaze, or \emph{implicit} like postures) are presented in an
homogeneous way.

The dialogue grounding process makes use of them at two distinct levels.

First, particular steps of the grounding process explicitly check for the
presence and value of specific facts: for instance, when a set of instances
match a category (the human says ``give me the bottle'' and the robot knows
about three bottles), the module may decide (it actually depends on the
quantifier preceding the class) to discard some of them based on their
\emph{visibility} by the speaker (implicit communication context built on the
human posture).

@@

Note that, while the system benefits from the complementary modalities, they
are not all required. The dialogue system would for instance run equally well
with only the verbal modality.

The second level of integration of multi-modality is implicit: by computing
symbolic properties from the geometry, richer descriptions and hence
discrimination possibilities are available: for instance, if reachability is
available, the robot may asks ``do you mean the bottle that is accessible to
me?'' to discriminate between the three bottles. That way, procedures relying
on discrimination transparently benefit from added modalities.

@@

Our execution controllers rely on symbolic task planning to convert faraway
desires into a succession of atomic actions. We use in our architecture the
HATP planner (for Human Aware Task Planner~\cite{Alili2009}).  HATP is based on
a Hierarchical Task Network (HTN) refinement, which performs an iterative task
decomposition into sub-tasks until reaching atomic actions.  The \emph{planning
domain} defines the set of methods describing how to decompose a task and can
be seen as the procedural knowledge of the robot.

@@

Second remark, after some research (see appendix B of~\cite{Lemaignan2012a} for
a detailed discussion)  we have decided not to represent neither the planning
domain nor the resulting plans in the knowledge base: the planning domain (with
tasks pre- and postconditions) is stored in a specific format, outside of the
knowledge repository, and the plans are directly communicated to the robot
controller. From our experience, the OWL formalism is not a satisfying
representation language for concepts that are both tightly bound to time
representation and non-monotonic.

@@

\emph{Experiences}, on the other hand, comprise of emotions, states and
questions (when asking a question, we consider the human to be in an
\emph{interrogative state}). When the knowledge base recognizes that an agent
\emph{experiences} a particular emotion or state, the execution controller may
decide to handle it, typically by trying to answer the question or using the
emotional or physical state as a parameter for subsequent actions. As an example,
when the speaker says ``I feel tired'', we change the motion planner
parametrization to lower the effort the human need to provide for the following
joint manipulation tasks. Note that this example has been implemented as a
proof-of-concept. We have not yet tried to define a theoretical framework that
would support action alteration based on the user's experienced states.

@@

The {\sc Oro} server proposes two paradigms to access its content: RPC-style
queries (based on the standard SPARQL language) or events. In its simplest
form, a module can register an event by passing through an event pattern, which
is basically a partial statement like \stmt{? type Elephant}, and a callback.
Every time a new instance of elephant appears in the knowledge base, the
callback is triggered.

@@

The robot controller designer does not need to directly care about how this
\emph{desire} is produced (this is delegated to perception modules), it can
focuses on the semantic of the desire.

Note also that we take advantage of the reasoning capabilities of the system:
for example, the goal of the action (\stmt{action hasGoal myself}) may well not
have been explicitly asserted, but inferred by the reasoner based on other
assertions.

@@

    \item{Knowledge is explicitly stored in one central and consistent
    repository of facts, accessible by all modules.}

@@

    \item{The first two points enable both a loosely-coupled architecture where
    modules can very easily be removed or replaced by other ones as long as
    they share the same semantics (modules are defined by the knowledge they
    produce),}

    \item{and a \emph{symbolic}, reactive, event-driven approach to robot control.
    By managing events at the same level as the reasoner, we take full
    advantage of the inference abilities of {\sc Oro} to trigger events whose
    \texttt{true} conditions can be inferred.}

@@

We believe that this \emph{knowledge-oriented} approach has a strong potential
not only to enable rich human-robot interaction, but also as a broader approach
to information alignment and fusion in complex robotic systems.  The
versatility of this paradigm could be illustrated by a simple imaginary
scenario with a blind robot and a deaf robot. The blind robot does not see (no
cameras or alike), but someone can verbally describe a scene to it. On the
other hand, the deaf robot has a good vision system, but cannot process verbal
input.  Without any changes to the software architecture that we described,
control modules of both robots could equally perform the same tasks since all
the knowledge is abstracted and centralised (note that to actually implement
this imaginary situation, the blind robot would of course need \textit{a
priori} 3D models of objects talked about to enable planning or pick and place
actions, and the deaf robot would require at least some gesture interpretation
to understand orders).

>>>>

\documentclass[preprint,12pt]{elsarticle}

\usepackage[english]{babel}
\usepackage{cite}
\usepackage{graphicx, subfigure}
\usepackage{url}
\graphicspath{{figs/}}
\DeclareGraphicsExtensions{.pdf,.jpg,.png}

@@

\newcommand{\stmt}[1]{{\footnotesize \tt $\langle$ #1\relax$\rangle$}}

@@

Over the last years we have focused our efforts on identifying the cognitive
prerequisites of these challenges, and giving them experimental reality on the
robots: what is required for sentences like ``Let's set the table together''
to be understood by the robot, correctly interpreted in the spatial and
temporal context of the interaction, and eventually transformed into a set of
actions.

We have chosen to tackle the challenge from several ends: human-aware
navigation and motion planning~\cite{Mainprice2011}, situation assessment
coupled with motion planning~\cite{Mainprice2012}, projection of
``mightabilities'' that anticipates what surrounding agents may
do~\cite{Pandey2011}, and the design and deployment of a pervasive
\emph{knowledge-oriented} software architecture.

The last item is the focus of this paper: what ``knowledge-oriented''
architecture means, and what its benefits and drawbacks are.

@@

In the next section, we first introduce the deliberative layer of our robots at
the functional level. It represents what we call a \emph{knowledge-oriented}
architecture, built around a central knowledge management tool called the
OpenRobot Ontology ({\sc Oro}) server.

The following sections present each of the facets of this deliberative layer:
how knowledge is produced asynchronously from geometric reasoning, how the
robot conducts grounded multi-modal interaction with humans, how knowledge is
used by decision-making components like the robot controller or the task
planner, and finally, how our knowledge-oriented architecture enables the
implementation of specific internal cognitive processes.

@@

\subsection{The LAAS deliberative layer}

Fig.~\ref{fig|archi} gives an overview of the connections between the
deliberative components of our architecture~\cite{Alami2011}.

@@

This architecture moves away from standard layered approaches. Interactions
between components at the deliberative level are mostly bidirectional and we do
not introduce layers of abstraction amongst software components\footnote{We do
have lower-level modules to execute actions or manage sensors, but all
cognition-related modules reside at the same level.}. This is nicely
illustrated by the dialogue input processing. This component does not simply
act as an alternative perceptual input to the symbolic database; it also
actively queries previously acquired knowledge to disambiguate and validate the
newly created symbolic knowledge (see section~\ref{sect|com}).

Our architecture relates to \emph{Beliefs, Desires, Intentions} (BDI)
architectures. BDI architectures are primarily focused on \emph{practical
reasoning}, \ie the process of deciding, step by step, which action to perform
to reach a goal (as summarized by Woolridge~\cite{Woolridge1999}). The
management of the interaction between knowledge (the beliefs) and task and plan
representation and execution (the desires and the intentions) is central, and
aims at selecting at each step the best subgoal. It becomes then an intention
that the robot commits to.

This interaction between knowledge and actions is also central to our approach
(as for any cognitive system): it is one of the activities of the
robot, actually shared between communication components (that can acquire
desires from interaction with agents, amongst other things) and an execution
controller that may decide to take an incoming desire into account to create
its own internal goals. The controller generates and manages intentions from
these goals with the help of a symbolic task planner, that also has direct
access to the knowledge base.

This activity is however not the backbone of our architecture. Other
activities are conducted in parallel, without being explicitly considered as
desires: assessment of the situation and the environment, dialogue (including
performative dialogue that can possibly change the internal state of the robot,
but does not lead to the creation of desires, like question answering or
statement assertion), various background monitoring and recognition tasks, etc.

Regarding the anchoring question, this architecture is bidirectional. The
components we described provide a \textit{bottom-up} grounding process:
geometric reasoning and dialogue processing modules constantly build and push
new symbolic contents about the world to the knowledge base where it becomes
accessible to decisional layers. In parallel, the knowledge base relies on
reasoning in a \textit{top-down} way to produce new facts that may in return
trigger physical behaviours.

@@

In our architecture (Fig.~\ref{fig|archi}), knowledge manipulation relies on a
\emph{semantic blackboard}: a central server (the {\sc Oro}
server~\cite{Lemaignan2010}) stores knowledge as it is produced by each of the
deliberative components. It conversely exposes a {\tt json}-based RPC API to
query the knowledge base~\cite{lemaignan2012kbapi}.

@@

Relying on RDF triples and Description Logics has advantages such as the
availability of numerous mature open-source libraries to manipulate the ontology,
interoperability with several major on-line knowledge bases (like {\sc
OpenCyc}, {\sc WordNet} or {\sc DBPedia}), open-world reasoning, and the formal
guarantee of decidability (it is always possible to classify a Description
Logics ontology).

It also has notable limitations, both fundamental (the suitability
of Description Logics when reasoning on --typically non-monotonic-- commonsense
knowledge is questionable) and practical: RDF triples imply only binary predicates
(\stmt{subject predicate object}), which constrains the expressiveness of the
system or leads to cumbersome reifications. Alternatives exist (like {\sc
KnowRob}~\cite{Tenorth2009a}) that mix RDF with more expressive logic
languages like {\sc Prolog}, at the price, however, of other limitations, like
closed-world reasoning or immutable T-Box. The classification
performance is another issue: from our experience, with an ontology sized for a
standard experiment (about 100 classes and 200 instances), classification
typically takes about 100ms, which becomes problematic during interactions.
Besides, the performances are difficult to predict, since a seemingly
inoffensive new statement may indirectly change radically the logical
complexity of the whole knowledge model and lead to notable degradation of
classification time.

This knowledge model also largely excludes representation of continuous
phenomena (like time) or uncertain phenomena. When required (for instance for
action recognition), these are managed inside the corresponding components, and
are not exposed at the semantic level.

Tools to manipulate and reason over ontologies are readily available, mature,
and already well accepted in the robotic community~\cite{Tenorth2009a,
Lim2011}. While alternatives like \emph{Answer Set Programming} have also been
successfully investigated in robotics~\cite{Chen2010,Erdem2012}, in particular
to deal with non-monotonic reasoning, we did not actually hit any brick wall
while working with OWL ontologies. We may reconsider this choice at a later
stage, but until now it has proven an effective framework to quickly explore
implementations of new cognitive abilities (for instance, it has been
conceptually and technically easy to add support for independent knowledge
models, one per the agents the robot interacts with --see
section~\ref{sect|tom}).

Besides, because ontologies and RDF statements are relatively simple concepts
to grasp, it also effectively helped to grow awareness amongst colleagues on
the significance of the ``semantic level'' when developing new components for
the robot.

@@

Theory of Mind (originally defined in~\cite{Premack1978}) is the cognitive
ability that a subject possesses to represent the mental state of another
agent, possibly including knowledge that contradicts the subject's own model: for
example, a book can be at the same time \emph{visible} for myself, and \emph{not
visible} for you.

Children develop this skill, which is essential to understand others' perspectives during
interactions, around the age of three. It supposes the
ability to build, store and retrieve separate models of the knowledge of the
interactors.

@@

One classical application of this cognitive skill is the so-called
\emph{False-Belief} experiment (also known as the \emph{Sally and Ann}
experiment)~\cite{Leslie2000}: a child is asked to watch a scene where two
people, A and B, manipulate objects. Then A leaves and B hides away one
object. When A comes back, we ask the child ``where do you think A will
look for the object?''. Before acquiring a theory of mind, children are not
able to separate their own (true) model of the world (where they know that
the object was hidden) from the model of A, which contains \emph{false
beliefs} on the world (A still thinks the object is at its original
position since he did not see B hiding it). Using separate knowledge models
in the knowledge base, we have been able to replicate this experience with
our robots~\cite{Warnier2012a}.

@@

The {\sc Oro} server also features a mechanism to mimic minimalistic forms of
biological memory.  When new statements are inserted in the knowledge base, a
\emph{memory profile} is optionally attached to them.

@@

We rely however on this short term memory for a particular use-case:
\emph{active concepts}. Some modules, like the natural language processor, use
the {\tt short term} memory profile to mark for a few seconds important
concepts that are currently manipulated by the robot. For example, if a human
asks the robot: ``Give me all red objects'', the human, the \concept{Give}
action, and each red objects that are found are successively marked as
\emph{active concepts} by inserting statements such as \stmt{human type
ActiveConcept} in the short-term memory (which can be considered, in this case,
to be a working memory). We use this feature to trace certain knowledge-related
processes.

@@

Anchoring perceptions in a symbolic model requires perception abilities and
their symbolic interpretation. We rely on a dedicated geometric and temporal
reasoning module called SPARK (\emph{SPAtial Reasoning \&
Knowledge}~\cite{Sisbot2011}). It is a situation assessment reasoner that
generates symbolic knowledge from the geometry of the environment with respect
to relations between objects, robots and humans (Fig.~\ref{fig|spark-oro}),
also taking into account the different perspective that each agent has on the
environment.

SPARK is an \emph{amodal} geometric model of the environment that serves both
as basis for the fusion of the perception modalities and as bridge with the
symbolic layer. This geometric model is continuously updated at run-time by the
robot based on its sensors (in our experiments, objects are identified and
localised through 2D barcodes, while humans are tracked with Kinect-like
devices, optionally assisted by motion capture to accurately track the head
motion, which is required to compute what the human is looking at).

@@

This allows us to deal with ambiguities that arise when one speaker refers to
an object within a reference system (or changes the reference system, \ie
switches perspective) without making the reference frame
explicit~\cite{Breazeal2006, Ros2010}.

As a result, the robot stores models of the environment either in the
\emph{ego-centric} reference frame (from the robot perspective) or in the
\emph{allo-centric} frame (addressee-centred).

@@

Natural language processing is one of the fields of human-robot interaction
for which the introduction of the semantic layer has been most beneficial.
In~\cite{Lemaignan2011a} we detail the techniques and the tool called
\texttt{dialogs} that we have developed for natural English language parsing and
grounding, along with verbalisation and (minimalist) dialogue management.

Natural language input (in experiments, we rely on an Android-based interface,
with Google speech recognition) is parsed into a grammatical structure, and
atoms of each sentence are resolved with the help of the ontology to ground
concepts like objects (\ie when a user says ``pick the can'', resolve to which
instance of \emph{can} the user is referring to) and actions. Sentences are
sorted into questions, desires and statements, and processed accordingly.

The system supports quantification (``give me \{a | the | some | all | any |
...\} can''), thematic roles (action-specific predicates that qualify the
actions), interactive disambiguation (the robot asks questions when it needs
more information), anaphora resolution (``give \emph{it} to me'') based on
dialogue history. It also permits knowledge extension by learning new semantic
structures (for instance, a sentence like ``learn that cats are animals'' is
converted into \stmt{Cat subClassOf Animal}), interprets common temporal and
place adverbs (like \emph{above} or \emph{tomorrow}) and translates to a
certain extend \emph{states} (``I'm tired'') into \emph{experiences}
(\stmt{HUMAN experiences state\_1, state\_1 hasFeature tired}).

@@

Besides, even knowledge that belongs to the \emph{common knowledge} may not be
appropriate in a given interaction context. For instance, the system may
compute that at a given instant the human is looking at the object: \stmt{human
looksAt obj1}. This property makes sense to both parties, but in the context of
the \emph{spygame}, we would like to mainly use immanent properties, not
volatile like a gaze. More research is required to identify relevant
interaction contexts and knowledge classes attached to them.

@@

Because all components rely on the same RDF formalism to format their outputs,
the different communication modalities (\emph{explicit} like verbal, deictic or
based on gaze, or \emph{implicit} like postures) are presented in a homogeneous
way. The dialogue grounding process makes use of them at two distinct levels.

First, particular steps of the grounding process explicitly check for the
presence and value of specific facts: for instance, when a set of instances
matches a category (the human says ``give me the bottle'' and the robot knows
about three bottles), the module may decide (it actually depends on the
quantifier preceding the class) to discard some of them based on their
\emph{visibility} for the speaker (implicit communication context built on the
human posture).

@@

Note that, while the system benefits from the complementary modalities, they
are not all required. The dialogue system would for instance happily run with
only the verbal modality, at the cost of weaker interaction.

The second level of integration of multi-modality is implicit: by computing
symbolic properties from the geometry, richer descriptions and hence
discrimination possibilities are available: for instance, if reachability is
available, the robot may ask ``do you mean the bottle that is accessible to
me?'' to discriminate between the three bottles. That way, procedures relying
on discrimination transparently benefit from added modalities.

@@

Our execution controllers rely on symbolic task planning to convert faraway
desires into a succession of atomic actions. We use in our architecture the
HATP planner (for Human Aware Task Planner~\cite{Alili2009}).  HATP is based on
a Hierarchical Task Network (HTN) refinement, which performs an iterative task
decomposition into sub-tasks until reaching atomic actions. The \emph{planning
domain} defines the set of methods describing how to decompose a task and
represents the procedural knowledge of the robot.

@@

Second remark, after some research (see appendix B of~\cite{Lemaignan2012a} for
a detailed discussion)  we have decided to represent neither the planning
domain nor the resulting plans in the knowledge base: the planning domain (with
task pre- and postconditions) is stored in a specific format, outside of the
central declarative knowledge repository, and the plans are directly
communicated to the robot controller. Thus, like many other cognitive
architectures, we have independent declarative and procedural knowledge stores.

@@

\emph{Experiences}, on the other hand, comprise of emotions, states and
questions (when asking a question, we consider the human to be in an
\emph{interrogative state}). When the knowledge base recognizes that an agent
\emph{experiences} a particular emotion or state, the execution controller may
decide to handle it, typically by trying to answer the question or using the
emotional or physical state as a parameter for subsequent actions. As an example,
when the speaker says ``I feel tired'', we change the motion planner
parametrization to lower the effort the human needs to provide for the following
joint manipulation tasks. Note that this example has been implemented as a
proof-of-concept. We have not yet tried to define a theoretical framework that
would support action alteration based on the user's experienced states.

@@

The {\sc Oro} server proposes two paradigms to access its content: RPC-style
queries (based on the standard SPARQL language) or events. In its simplest
form, a module can subscribe to an event by passing through an event pattern
(in its basic form, a partial statement like \stmt{? type Elephant}) and a
callback.  Each time a new instance of elephant appears in the knowledge base,
the callback is triggered.

@@

The robot controller designer does not need to directly care about how this
\emph{desire} is produced (this is delegated to perception modules), he can
focus on the semantic of the desire.

Note also that we take advantage of the reasoning capabilities of the system:
for example, the goal of the action (\stmt{action hasGoal myself}) may not be
explicitly asserted, but inferred by the reasoner based on other assertions.

@@

    \item{Knowledge is explicitly stored in one central and consistent
    repository of facts, accessible for all modules.}

@@

    \item{The first two points enable a loosely-coupled architecture where
    modules can be removed or replaced easily by other ones as long as
    they share the same semantics (modules are defined by the knowledge they
    produce),}

    \item{We adopt a \emph{symbolic}, reactive, event-driven approach to robot control.
    By managing events at the same level as the reasoner, we take full
    advantage of the inference abilities of {\sc Oro} to trigger events whose
    \texttt{true} conditions can be inferred.}

@@

Those items are not new \emph{per se}. It seems however interesting to
underline the shift of focus this implies during the design and integration
phases. The
versatility of this paradigm could be illustrated by a simple imaginary
scenario with a blind robot and a deaf robot. The blind robot does not see (no
cameras or alike), but someone can verbally describe a scene to it. On the
other hand, the deaf robot has a good vision system, but cannot process verbal
input.  Without any changes to the software architecture that we described,
control modules of both robots could equally perform the same tasks since all
the knowledge is abstracted and centralised (note that to actually implement
this imaginary situation, the blind robot would of course need \textit{a
priori} 3D models of objects talked about to enable planning or pick and place
actions, and the deaf robot would require at least some gesture interpretation
to understand orders).
